---
title: 'DATA 605:  Week 3 Assignment'
author: "Aaron Grzasko"
date: "February 19, 2017"
output: 
    html_document:
        theme: default
        highlight: haddock
---  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1  
  
**1.  What is the rank of matrix A?**  
  
$$A=\begin{bmatrix} 1 & 2 & 3 & 4 \\ -1 & 0 & 1 & 3 \\ 0 & 1 & -2 & 1 \\ 5 & 4 & -2 & -3 \end{bmatrix}$$  
The rank of a matrix is equal to the number of pivots.  
  
To determine the number of pivots, we perform row operations until the matrix is in upper triangular form, U.  

We will now perform the row operations manually, using elimination matrices:  

```{r}
A <- matrix(c(1,2,3,4,-1,0,1,3,0,1,-2,1,5,4,-2,-3),nrow=4,ncol=4,byrow=TRUE)
U <- A

# row2 = row1 + row2
E21 <- diag(4); E21[2,1] <- 1
U <- E21 %*% A
```
$$\sim \begin{bmatrix} 1 & 2 & 3 & 4 \\ 0 & 2 & 4 & 7 \\ 0 & 1 & -2 & 1 \\ 5 & 4 & -2 & -3 \end{bmatrix}$$
```{r}
# row4 = -5*row1 + row4 
E41 <- diag(4); E41[4,1] <- -5 
U <- E41 %*% U
```
$$\sim \begin{bmatrix} 1 & 2 & 3 & 4 \\ 0 & 2 & 4 & 7 \\ 0 & 1 & -2 & 1 \\ 0 & -6 & -17 & -23 \end{bmatrix}$$
```{r}
# row3 = row2 - 2*row3
E32 <- diag(4); E32[3,2] <- 1; E32[3,3] <- -2 
U <- E32 %*% U
```
$$\sim \begin{bmatrix} 1 & 2 & 3 & 4 \\ 0 & 2 & 4 & 7 \\ 0 & 0 & 8 & 5 \\ 0 & -6 & -17 & -23 \end{bmatrix}$$
```{r}
# row4 = 3*row2 + row4
E42 <- diag(4); E42[4,2] <- 3 
U <- E42 %*% U
```
$$\sim \begin{bmatrix} 1 & 2 & 3 & 4 \\ 0 & 2 & 4 & 7 \\ 0 & 0 & 8 & 5 \\ 0 & 0 & -5 & -2 \end{bmatrix}$$
```{r}
# row4 = 5/8*row3 + row4
E43 <- diag(4); E43[4,3] <- 5/8
U <- E43 %*% U 

# print U
U
```
$$\sim \begin{bmatrix} 1 & 2 & 3 & 4 \\ 0 & 2 & 4 & 7 \\ 0 & 0 & 8 & 5 \\ 0 & 0 & 0 & 9/8 \end{bmatrix}$$  
There are four pivots in the row echelon matrix, U; therefore, the <span style = "color: red">rank of matrix A is 4</span>.  

Now let's check our work against the `rankMatrix()` function from the Matrix package:  
```{r}
# load Matrix package, if necessary
if (!require(Matrix)){install.packages('Matrix');require(Matrix)}

# calculate rank of matrix A
rankMatrix(A)[1]
```

**2. Given an m x n matrix where m > n, what can be the maximum rank? The minimum
rank, assuming that the matrix is non-zero?**  
  
The rank of a matrix can be is no larger than *min(m,n)*.  In this example, n is less than m; so the <span style = "color: red">maximum rank is n</span>.  
  
The minimum rank is zero only for zero matrices.  For all other matrices, the minimum rank is one. Because we assume our matrix is non-zero, the <span style = "color: red">minimum rank is 1</span>.  
  
**3. What is the rank of matrix B?**  
  
$$B=\begin{bmatrix} 1 & 2 & 1 \\ 3 & 6 & 3 \\ 2 & 4 & 2 \end{bmatrix}$$  
We notice that that row 2 is equal to row 1 times 3, and row 3 is equal to row 1 times two.  Similarly, column 2 is equal to column 1 times 2, and column 3 is identical to column 1. In other words, rows 2 and 3 are not linearly independent from row 1; and columns 2 and 3 are not linearly independent from column 1.  
  
When we perform subsequent row operations to obtain row echelon form, we expect only one row comprising at least one, nonzero element.  That is, we expect only one pivot; meaning the <span style = "color: red">rank of Matrix B is 1</span>.  
  
Let's now verify by applying row reduction procedures:  
  
```{r}
B <- matrix(c(1,2,1,3,6,3,2,4,2),nrow=3,ncol=3,byrow=TRUE)

U <- B

# row2 = -3*row1 + row2
E21 <- diag(3); E21[2,1] <- -3
U <- E21 %*% U
```
$$\sim \begin{bmatrix} 1 & 2 & 1 \\ 0 & 0 & 0 \\ 2 & 4 & 2 \end{bmatrix}$$
```{r}
# row3 = -2*row1 + row3
E31 <- diag(3); E31[3,1] <- -2
U <- E31 %*% U

# print u
U
```
$$\sim \begin{bmatrix} 1 & 2 & 1 \\ 0 & 0 & 0 \\ 0 & 0 & 0 \end{bmatrix}$$
We have just verified that the row-echelon matrix has only one pivot; so the <span style = "color: red">rank of Matrix B is 1</span>.  
  
## Problem Set 2  
  
**1. Compute the eigenvalues and eigenvectors of the matrix A. You’ll need to show your work. You’ll need to write out the characteristic polynomial and show your solution.**  
  
$$A=\begin{bmatrix} 1 & 2 & 3 \\ 0 & 4 & 5 \\ 0 & 0 & 6 \end{bmatrix}$$
To find the eigenvalues or matrix A, we solve for $\lambda$ in the following equation:  
$$A\overrightarrow { v } =\lambda \overrightarrow { v }$$  
$$\lambda{ I }_{ 3 } \overrightarrow { v }-A\overrightarrow { v }=\overrightarrow { 0 }$$
$$(\lambda { I }_{ 3 }-A)\overrightarrow { v } =\overrightarrow { 0 } $$  
The equation above is true for nonzero vector $\overrightarrow { v }$  if and only if:  
$$det(\lambda { I }_{ 3 }-A)=0$$ 
Now, let's substitute the matrix A and 3 x 3 identity matrix into the equation:  

$$det(\begin{bmatrix} \lambda  & 0 & 0 \\ 0 & \lambda  & 0 \\ 0 & 0 & \lambda  \end{bmatrix}-\begin{bmatrix} 1 & 2 & 3 \\ 0 & 4 & 5 \\ 0 & 0 & 6 \end{bmatrix})=0$$ 
$$det(\begin{bmatrix} \lambda -1 & -2 & -3 \\ 0 & \lambda -4 & -5 \\ 0 & 0 & \lambda -6 \end{bmatrix})=0$$  

Apply cofactor expansion down the first column of the matrix above:  

$${ (-1) }^{ 1+1 }(\lambda -1)det(\begin{bmatrix} \lambda -4 & -5 \\ 0 & \lambda -6 \end{bmatrix})=0$$
$$(\lambda -1)\left[ (\lambda -4)(\lambda -6)-(-5)*0 \right] =0$$
$$(\lambda -1)({ \lambda  }^{ 2 }-10\lambda +24)=0$$
<span style = "color: red">This produces the characteristic polynomial equation:  
$$({ \lambda  }^{ 3 }-11{ { \lambda  }^{ 2 } }+34\lambda -24)=0$$</span>  
The equation can now be factorized as follows:  
$$(\lambda -1)(\lambda -6)(\lambda -4)=0$$
<span style = "color: red">The eigenvalues of A are as follows:</span>  

<span style = "color: red">$$\lambda=1, \lambda=6, \lambda=4$$</span>
  
We now substitute each eigenvalue into the equation, $(\lambda{ I }_{ 3 }-A)\overrightarrow { v }=\overrightarrow { 0 }$  

For $\lambda=1$:  
$$\begin{bmatrix} 0 & -2 & -3 \\ 0 & -3 & -5 \\ 0 & 0 & -5 \end{bmatrix}\begin{bmatrix} { v }_{ 1 } \\ { v }_{ 2 } \\ { v }_{ 3 } \end{bmatrix}=\begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}$$  
Using back substitution, we solve as follows:
$${ v }_{ 1 } = free$$
$${ v }_{ 2 } = 0$$
$${ v }_{ 3 } = 0$$
<span style = "color: red">$${ E }_{ \lambda =1 }={span( }\left[ \begin{matrix} 1 \\ 0 \\ 0 \end{matrix} \right] )$$</span>


  
For $\lambda=6$:  
$$\begin{bmatrix} 5 & -2 & -3 \\ 0 & 2 & -5 \\ 0 & 0 & 0 \end{bmatrix}\begin{bmatrix} { v }_{ 1 } \\ { v }_{ 2 } \\ { v }_{ 3 } \end{bmatrix}=\begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}$$  
Using back substitution we solve as follows:  
$${ v }_{ 1 } =\frac { 2 }{ 5 } { v }_{ 2 }+\frac { 3 }{ 5 } { v }_{ 3 }=\frac { 2 }{ 5 } \frac { 5 }{ 2 } { v }_{ 3 }+\frac { 3 }{ 5 } { v }_{ 3 }=\frac { 8 }{ 5 } { v }_{ 3 } $$
$${ v }_{ 2 } = \frac { 5 }{ 2 } { v}_{ 3 }$$


$${ v }_{ 3 } = free$$  
$${ E }_{ \lambda =6 }={span( }\left[ \begin{matrix}\frac { 8 }{ 5 }  \\ \frac { 5 }{ 2 }  \\ 1 \end{matrix} \right]) $$
It is customary to define the eigenspace using unit vectors.  
First calculate the length of the vector above:  
$$\sqrt { { \left( \frac { 8 }{ 5 }  \right)  }^{ 2 }+{ \left( \frac { 5 }{ 2 }  \right)  }^{ 2 }+{ 1 }^{ 2 } } =\quad 3.1321$$  
Now, divide each element of the vector calculated above by the length:  
<span style = "color: red">$${ E }_{ \lambda =6 }=span(\left[ \begin{matrix} 0.5108 \\ 0.7982 \\ 0.3193 \end{matrix} \right] )$$</span>  

  
For $\lambda=6$:  
$$\begin{bmatrix} 3 & -2 & -3 \\ 0 & 0 & -5 \\ 0 & 0 & -2 \end{bmatrix}\begin{bmatrix} { v }_{ 1 } \\ { v }_{ 2 } \\ { v }_{ 3 } \end{bmatrix}=\begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}$$ 
Using back substitution, we solve as follows:
$${ v }_{ 1 } =\frac { 2 }{ 3 } { v }_{ 3 } $$  
$${ v }_{ 2 } = free$$  
$${ v }_{ 3 } = 0$$
$${ E }_{ \lambda =4 }={span( }\left[ \begin{matrix}\frac { 2 }{ 3 }  \\ 1\\ 0 \end{matrix} \right]) $$

To convert to a unit vector, calculate the length of the vector above: 
$$\sqrt { { \left( \frac { 2 }{ 3 }  \right)  }^{ 2 }+{ 1 }^{ 2 }+{ 0 }^{ 2 } } =1.2019$$
Now divide each element of the the vector above by the length:  
<span style = "color: red">$${ E }_{ \lambda =4 }={span( }\left[ \begin{matrix}0.5547  \\ 0.8321\\ 0 \end{matrix} \right]) $$</span>  
  
To summarize, our calculated eigenvectors for $\lambda=1$, $\lambda=6$, and $\lambda=4$, respectively, are:  
<span style = "color: red">$$\left[ \begin{matrix} 1 \\ 0 \\ 0 \end{matrix} \right] ,\left[ \begin{matrix} 0.5108 \\ 0.7982 \\ 0.3193 \end{matrix} \right] ,\quad and\quad \left[ \begin{matrix} 0.5547 \\ 0.8321 \\ 0 \end{matrix} \right] $$</span>  

Finally, let's verify our calculated eigenvalues and eigenvectors using the built-in R function, `eigen()`:  
```{r}
A <- matrix(c(1,2,3,0,4,5,0,0,6),3,3,byrow=TRUE)  
eigen(A)

```  
  





